Unlocking Sensitivity for Visibility-based Estimators of the 21\,cm Reionization Power Spectrum

### 2017-07-22 Rotation paper/draft_save.tex
incorporated Aaron's last round comments

A crucial concern when measuring the $21\,\textrm{cm}$ power spectrum is the possibility of astrophysical foreground contamination. Because foregrounds are orders of magnitude larger than the cosmological signal, even small residuals from an imperfect isolation of these foregrounds can masquerade as a false detection. Recent literature treatments of this problem have shown that if one combines the mathematical properties of interferometric measurement with the empirical fact that foregrounds are spectrally smooth, a clean separation of foregrounds and cosmological signal can be achieved in harmonic space. These studies suggest that foregrounds preferentially appear in harmonic modes of the sky corresponding to fluctuations that are angularly fine but spectrally smooth, a region in harmonic space colloquially known as ``the wedge" \citep{Datta2010,Vedantham2012,Morales2012,delay-transform,Trott2012,Thyagarajan2013,pober_et_al2013b,dillon_et_al2014,Hazelton2013,Thyagarajan_et_al2015a,Thyagarajan_et_al2015b,wedge1, wedge2,chapman_et_al2016,pober_et_al2016,seo_and_hirata2016,jensen_et_al2016,kohn_et_al2016}. A foreground mitigation strategy can then consist mainly of avoiding those modes by working in the complementary region known as ``the EoR window". However, for this strategy to be viable, it is necessary to show that one's data analysis pipeline does not corrupt the separation between the wedge and the EoR window. We will now do so for our proposed two-baseline power spectrum estimator.

@@

Traditionally, spectral and angular dimensions are cylindrically binned with the coordinates $k_\bot$ and  $k_\parallel$, where the angular direction $k_\bot$ is probed by the beam and line-of-sight direction $k_\parallel$ is probed by the frequency spectrum. Despite their usefulness, however, $(k_\bot, k_\parallel)$ are local coordinates that are well defined only in the narrow field-of-view limit. 
As discussed in previous sections, the combination of information from nearly equivalent baselines for wide-field telescopes is inherently a three-dimensional problem. Although the estimator of Section \ref{sec:method} does not have such limitations, to discuss foreground signatures with possible curved sky effects, we must employ a set of basis functions where angular fluctuations are encoded by spherical harmonics, and radial fluctuations by spherical Bessel functions.

\cite{sphere} discusses in detail power-spectrum analyses using spherical harmonics. In such as basis, modes in the sky are indexed by the magnitude $k$ of the wavevector $\mathbf{k}$ and spherical harmonic indices $(\ell, m)$. In other words, one defines
\begin{equation}
\overline{T}_{\ell m} (k) \equiv \sqrt{\frac{2}{\pi}} \int d\Omega dr r^2 j_\ell (kr) Y_{\ell m}^* (\hat{\mathbf{r}}) T (\hat{\mathbf{r}}),
\end{equation}
where $Y_{\ell m}$ is a spherical harmonic function and $j_\ell$ is the $\ell$th-order spherical Bessel function of the first kind. The power spectrum is then related to these spherical harmonic Bessel modes via the relation
\begin{equation}
\label{eq:SHB}
\langle \overline{T}_{\ell m} (k) \overline{T}_{\ell^\prime m^\prime}^* (k^\prime) \rangle = \frac{\delta^D (k - k^\prime)}{k^2} \delta_{\ell \ell^\prime} \delta_{m m^\prime} P(k).
\end{equation}

A delay-transformed visibility $V(\mathbf{b}, \tau) $ is related to the spherical Fourier-Bessel modes by
\begin{equation}
V(\mathbf{b}, \tau) = \sqrt{\frac{2}{\pi}} \sum_{\ell m} \int \! dk\, k^2 g_{\ell m} (k; \mathbf{b}, \tau) \overline{T}_{\ell m} (k),
\end{equation}
where
\begin{equation}
g_{\ell m} (k ; \mathbf{b}, \tau) = \int \! d\Omega d\nu B(\hat{\mathbf{r}},\nu) Y_{\ell m} (\hat{\mathbf{r}}) j_\ell (k r) e^{i 2 \pi \nu (\tau - \mathbf{b}\cdot \hat{\mathbf{r}} / c)}
\end{equation}
and the analogous quantity for $V_\psi$ is given by
\begin{eqnarray}
\label{eq:glmpsi}
g_{\ell m}^\psi (k ; \mathbf{b}, \tau) = \int && d\Omega d\nu B(\boldsymbol \Gamma \hat{\mathbf{r}},\nu) Y_{\ell m} (\hat{\mathbf{r}}) j_\ell (k r) \nonumber \\
&& \times  e^{i 2 \pi \nu (\tau - \mathbf{b}\cdot \boldsymbol \Gamma \hat{\mathbf{r}} / c) + i \psi_\nu}.
\end{eqnarray}
Note that unlike the primary beam and the fringe pattern, the spherical harmonic function does \emph{not} rotate, since it originated from a spherical harmonic expansion of the sky temperature, which is fixed. With these expressions and a little algebra combining Eqs. \eqref{eq:SHB} through \eqref{eq:glmpsi}, we arrive at a form of the two-baseline estimator (compare with Eq. \eqref{eq:final}):
\begin{eqnarray}
\hat{P}(k) &=& \frac{ \langle V^{*}(\boldsymbol{b},\tau)V_{\psi}(\boldsymbol{b'},\tau) \rangle}{\Theta} \nonumber \\
&=& \sum_\ell \int \!dk\, W_{\ell}(k; \mathbf{b}, \mathbf{b}^\prime , \tau) P(k),
\end{eqnarray}
where we have defined the \emph{window function}
\begin{equation}
W_{\ell}(k; \mathbf{b}, \mathbf{b}^\prime , \tau) \equiv \frac{2 k^2}{\pi \Theta} \sum_m g_{\ell m} (k ; \mathbf{b}, \tau) g_{\ell m}^{\psi*} (k ; \mathbf{b}^\prime, \tau).
\end{equation}
By construction, the window function sums to unity when integrated over all $k$ and summed over all $\ell$. We may therefore interpret our estimator $\hat{P}$ of the power spectrum to be a weighted average over all possible $\ell$ and $k$ modes, with the window function providing the weights of this average. This means that window functions can serve as indicators of foreground leakage into the EoR window. One writes down the estimator for a hypothetical measurement of a power spectrum mode within the EoR window, and additionally computes the window function for that estimator. Ideally, the window function for $\hat{P}(k)$ will be sharply peaked around $k$. In general, however, window functions will have wings that encroach upon other modes. If the window function is substantially non-zero at low-$k$ modes---which is where foreground emission typically resides, given its spectral smoothness---then the EoR window will be contaminated by foregrounds.

Figure \ref{fig:multiblwindows} shows four example window functions. The two window functions in the top plot are for equivalent baseline power-spectrum estimators, i.e., where one squares the delay spectrum from a single baseline and normalizes the result. The window function localized at low $\ell$ is for the baseline \{2,0\} of around $30\,\textrm{m}$ in length, while the window function localized at high $\ell$ is for \{12,0\}, roughly $180\,\textrm{m}$ in length. The window functions for near-equivalent baseline estimates are shown in the bottom plot, for \{2,0:2,1\} and \{12,0:12,1\} at low- and high-$\bar{b}$, respectively. In all cases, the window functions that we show are for $\tau = 703\,\textrm{ns}$, and are generated using primary beam models for PAPER. For numerical convenience we assumed that the bandpass $\phi_\nu$ (see Eq. \eqref{eq:B}) takes the form of a half-period sine curve that peaks at $150\,\textrm{\,MHz}$ and goes to zero $5\,\textrm{\,MHz}$ on either side of the peak.

>>>>

A crucial concern when measuring the $21\,\textrm{cm}$ power spectrum is the possibility of astrophysical foreground contamination. Being orders of magnitude larger than the cosmological signal, even small residuals from an imperfect isolation of these foregrounds can masquerade as a false detection. Recent literature treatments of this problem have shown that if one combines the mathematical properties of interferometric measurement with the empirical fact that foregrounds are spectrally smooth, a clean separation of foregrounds and cosmological signal can in principle be achieved in harmonic space \citep{wedge1, wedge2}. These studies suggest that foregrounds preferentially appear in harmonic modes of the sky corresponding to fluctuations that are angularly fine but spectrally smooth, a region in harmonic space colloquially known as ``the wedge". One's foreground mitigation strategy can then consist mainly of avoiding those modes by working in the complementary region known as ``the EoR window". However, for this strategy to be viable, it is necessary to show that one's data analysis pipeline does not corrupt the separation between the wedge and the EoR window. We will now do so for our proposed two-baseline power spectrum estimator.

@@

Traditionally, spectral and angular dimensions are cylindrically binned with the coordinates $k_\bot$ and  $k_\parallel$, where the angular direction $k_\bot$ is probed by the beam and line-of-sight direction $k_\parallel$ is probed by the frequency spectrum. Despite their usefulness, however, $(k_\bot, k_\parallel)$ are local coordinates that are well defined only in the narrow field-of-view limit. 
As discussed in previous sections, the combination of information from nearly equivalent baselines for wide-field telescopes is inherently a three-dimensional problem. Although the estimator of Section \ref{sec:method} does not have such limitations, to discuss foreground signature with possible curved sky effects, we must employ a set of basis functions where angular fluctuations are encoded by spherical harmonics, and radial fluctuations by spherical Bessel functions.

\cite{sphere} discusses in detail power spectrum analyses using spherical harmonics. In such as basis, modes in the sky are indexed by the magnitude $k$ of the wavevector $\mathbf{k}$ and spherical harmonic indices $(\ell, m)$. In other words, one defines
\begin{equation}
\overline{T}_{\ell m} (k) \equiv \sqrt{\frac{2}{\pi}} \int d\Omega dr r^2 j_\ell (kr) Y_{\ell m}^* (\hat{\mathbf{r}}) T (\hat{\mathbf{r}}),
\end{equation}
where $Y_{\ell m}$ is a spherical harmonic function and $j_\ell$ is the $\ell$th-order spherical Bessel function of the first kind. The power spectrum is then related to these spherical harmonic Bessel modes via the relation
\begin{equation}
\label{eq:SHB}
\langle \overline{T}_{\ell m} (k) \overline{T}_{\ell^\prime m^\prime}^* (k^\prime) \rangle = \frac{\delta^D (k - k^\prime)}{k^2} \delta_{\ell \ell^\prime} \delta_{m m^\prime} P(k).
\end{equation}

A delay-transformed visibility $V(\mathbf{b}, \tau) $ is related to the spherical Fourier-Bessel modes by
\begin{equation}
V(\mathbf{b}, \tau) = \sqrt{\frac{2}{\pi}} \sum_{\ell m} \int \! dk\, k^2 g_{\ell m} (k; \mathbf{b}, \tau) \overline{T}_{\ell m} (k),
\end{equation}
where
\begin{equation}
g_{\ell m} (k ; \mathbf{b}, \tau) = \int \! d\Omega d\nu B(\hat{\mathbf{r}},\nu) Y_{\ell m} (\hat{\mathbf{r}}) j_\ell (k r) e^{i 2 \pi \nu (\tau - \mathbf{b}\cdot \hat{\mathbf{r}} / c)}
\end{equation}
and the analogous quantity for $V_\psi$ is given by
\begin{eqnarray}
\label{eq:glmpsi}
g_{\ell m}^\psi (k ; \mathbf{b}, \tau) = \int && d\Omega d\nu B(\boldsymbol \Gamma \hat{\mathbf{r}},\nu) Y_{\ell m} (\hat{\mathbf{r}}) j_\ell (k r) \nonumber \\
&& \times  e^{i 2 \pi \nu (\tau - \mathbf{b}\cdot \boldsymbol \Gamma \hat{\mathbf{r}} / c) + i \psi_\nu}.
\end{eqnarray}
Note that unlike the primary beam and the fringe pattern, the spherical harmonic function does \emph{not} rotate, since it originated from a spherical harmonic expansion of the sky temperature, which is fixed. With these expressions and a little algebra combining Eqs. \eqref{eq:SHB} through \eqref{eq:glmpsi}, we arrive at a form of the two-baseline estimator (compare with Eq. \eqref{eq:final}):
\begin{eqnarray}
\hat{P}(k) &=& \frac{ \langle V^{*}(\boldsymbol{b},\tau)V_{\psi}(\boldsymbol{b'},\tau) \rangle}{\Theta} \nonumber \\
&=& \sum_\ell \int \!dk\, W_{\ell}(k; \mathbf{b}, \mathbf{b}^\prime , \tau) P(k),
\end{eqnarray}
where we have defined the \emph{window function}
\begin{equation}
W_{\ell}(k; \mathbf{b}, \mathbf{b}^\prime , \tau) = \frac{2 k^2}{\pi \Theta} \sum_m g_{\ell m} (k ; \mathbf{b}, \tau) g_{\ell m}^{\psi*} (k ; \mathbf{b}^\prime, \tau).
\end{equation}
By construction, the window function sums to unity when integrated over all $k$ and summed over all $\ell$. We may therefore interpret our estimator $\hat{P}$ of the power spectrum to be a weighted average over all possible $\ell$ and $k$ modes, with the window function providing the weights of this average. This means that window functions can serve as indicators of foreground leakage into the EoR window. One writes down the estimator for a hypothetical measurement of a power spectrum mode within the EoR window, and additionally computes the window function for that estimator. Ideally, the window function for $\hat{P}(k)$ will be sharply peaked around $k$. In general, however, window functions will have wings that encroach upon other modes. If the window function is substantially non-zero at low-$k$ modes---which is where foreground emission typically resides, given its spectral smoothness---then the EoR window will be contaminated by foregrounds.

Figure \ref{fig:multiblwindows} shows four example window functions. The two window functions in the top plot are for equivalent baseline power spectrum estimators, i.e., where one squares the delay spectrum from a single baseline and normalizes the result. The window function localized at low $\ell$ is for the baseline \{2,0\} of around $30\,\textrm{m}$ in length, while the window function localized at high $\ell$ is for \{12,0\}, roughly $180\,\textrm{m}$ in length. The window functions for near-equivalent baseline estimates are shown in the bottom plot, for \{2,0:2,1\} and \{12,0:12,1\} at low- and high-$\bar{b}$, respectively. In all cases, the window functions that we show are for $\tau = 703\,\textrm{ns}$, and are generated using primary beam models for PAPER. For numerical convenience we assumed that the bandpass $\phi_\nu$ (see Eq. \eqref{eq:B}) takes the form of a half-period sine curve that peaks at $150\,\textrm{\,MHz}$ and goes to zero $5\,\textrm{\,MHz}$ on either side of the peak.
